---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.path = "man/figures/README-")
```

# SporeLag

<!-- badges: start -->
[![R-CMD-check](https://github.com/friveramariani/SporeLag/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/friveramariani/SporeLag/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/friveramariani/SporeLag/graph/badge.svg)](https://app.codecov.io/gh/friveramariani/SporeLag)
[![DOI](https://zenodo.org/badge/1299833349.svg)](https://doi.org/10.5281/zenodo.21364422)
<!-- badges: end -->

Deterministic, group-safe utilities that turn daily environmental exposure
series — pollen and spore counts, and other time-varying exposures such as
ozone or particulate matter — into analysis-ready lagged and moving-average
features for environmental epidemiology.

## Why

Building a lagged exposure looks like a one-liner. It isn't, because daily
surveillance data has two different defects that are easy to conflate:

- a **missing value** is a hole in a column — `NA` where a sample failed;
- a **gap** is a hole in *time* — a day with no row at all.

`is.na()` cannot see a gap. And a lag computed by row position on a gapped
series is a lag by *position*, not by *time*: a "1-day lag" silently becomes
however many days happen to separate two adjacent rows. The resulting exposure
column looks plausible, runs cleanly through a regression, and yields a
misaligned lag–response estimate that nothing downstream will flag.

SporeLag makes that failure impossible: the temporal functions refuse to
operate on a gapped grid.

## Installation

```r
# install.packages("remotes")
remotes::install_github("<user>/SporeLag")
```

## Usage

```{r}
library(SporeLag)

model_ready <- pollen_demo |>
  complete_daily_grid(date = "date", by = "site") |>   # close gaps in TIME
  assign_iso_week(date = "date") |>                    # iso_year + iso_week
  assign_season(date = "date") |>                      # season label
  impute_weekly_mean(value = "count", by = "site") |>  # fill NAs, flag them
  build_moving_average(value = "count_imputed", window = c(3, 7),
                       date = "date", by = "site") |>
  apply_lag(value = "count_imputed", lags = 0:3,
            date = "date", by = "site")

names(model_ready)
```

`complete_daily_grid()` must come first, or the last two steps error. The
pipeline enforces its own correctness.

## Functions

| Function | Purpose | Appends |
|---|---|---|
| `complete_daily_grid()` | insert rows for absent calendar days | rows, not columns |
| `assign_iso_week()` | ISO 8601 week **and** year | `iso_week`, `iso_year` |
| `assign_season()` | configurable season label | `season` |
| `impute_weekly_mean()` | fill `NA`s with the ISO-week mean | `{value}_imputed`, `{value}_imputed_flag` |
| `build_moving_average()` | trailing windowed mean | `{value}_ma{k}` |
| `apply_lag()` | time-indexed lags | `{value}_lag{n}` |

## Design principles

- **Append, never modify.** A data frame goes in; the same data frame comes out
  with new columns. Input columns are never dropped, reordered, or overwritten.
- **Deterministic.** No randomness. Identical input, identical output — on every
  operating system.
- **Group-safe.** `by` is a character vector of column names. Values never bleed
  across a group boundary.
- **Fail loudly.** Gaps, duplicate dates, `NA` group keys, and negative lags all
  raise classed errors rather than producing plausible-looking wrong answers.
- **Minimal dependencies.** `rlang` and `cli`. That's it.

## Scope

SporeLag builds exposure variables. It does not fit models, merge outcomes, or
define pollen seasons for you — those are analytic decisions that belong to you,
not to a preprocessing package.

## Learn more

`vignette("getting-started", package = "SporeLag")` walks through the pipeline
and, more importantly, through the ways it can go wrong.
